The truncated KdV framework for modeling anomalous waves induced by abrupt depth changes

### 2019-07-15 WavesData-2019/LatexNotes/notes.tex
NM

% Basic physical parameters and scales
\newcommand{\depth}{h}
\newcommand{\dup}{\depth_{-}}
\newcommand{\ddn}{\depth_{+}}
\newcommand{\freqp}{f_p}
\newcommand{\lam}{\lambda}
\newcommand{\lamup}{\lam_{-}}
\newcommand{\lamfac}{N}
% More parameters
\newcommand{\etastd}{\eta_{\text std}}
\newcommand{\amp}{a}
% Statistical quantities
\newcommand{\skw}{\text{skew}}
\newcommand{\var}{\text{var}}
% Dimensionless parameters
\newcommand{\epsup}{\eps_0}
\newcommand{\delup}{\delta_0}
\newcommand{\drat}{D}
\newcommand{\ampp}{\mathcal{A}}
% Hamiltonian and Gibbs stuff
\newcommand{\Ham}{H}
\newcommand{\Hp}{\Ham^{+}}
\newcommand{\Hm}{\Ham^{-}}
\newcommand{\Gibbs}{\mathcal{G}}
\newcommand{\Gp}{\Gibbs^{+}}
\newcommand{\Gm}{\Gibbs^{-}}
\newcommand{\meanup}[1]{\mean{#1}_{-}}
\newcommand{\meandn}[1]{\mean{#1}_{+}}

@@

%\author{
% Tyler Bolles
%Andrew J.~Majda\thanks{Courant Institute of Mathematical Sciences}, 
%M.~N.~J.~Moore\thanks{Florida State University}, 
%Di Qi\thanks{Courant Institute of Mathematical Sciences} }
\maketitle

@@

We will primarily consider the case in which waves originate from a region of constant depth, encounter an abrupt depth change, and then continue in another region of constant depth. Thus, depth will be piecewise constant
\begin{align}
\depth = 
\begin{cases}
\it \dup \quad \mbox{if } x<0 \\
\it \ddn \quad \mbox{if } x>0
\end{cases}
\end{align}
Most often, we will consider waves moving into shallower depth, so that $\dup > \ddn$. The incoming waves are randomized and generated with a peak forcing frequency of $\freqp$, which gives rise to peak wavelength $\lam = c/\freqp = \sqrt{g \depth} / \freqp$.

@@

% Table
%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
\begin{table}[h]%[htbp]
\begin{center}
\caption{Table of parameters}
\label{paramtable}
\begin{tabular}{l l l}
\hline \multicolumn{3} { c }{Parameters that are constant in a single experiment} \\
\hline Description & Notation and definition & Value in experiments \\
\hline
Peak forcing frequency		& $f_p$						& 2 Hz \\
Upstream depth			& $\dup$						& 12.5 cm \\
Characteristic wave amplitude	& $a = \mean{\eta^2}^{1/2} $		& 0.03--0.3 cm \\
Upstream wavelength		& $\lamup = \sqrt{g \dup}/f_p$		& 56 cm \\
Amplitude-to-depth ratio		& $\epsup = a / \dup$			& 0.0024--0.024 \\
Depth-to-wavelength ratio		& $\delup = \dup / \lamup$		& 0.22 \\
\hline \multicolumn{3} { c }{Parameters that change value across ADC} \\
\hline Description & Notation and definition \\
\hline
Local depth			& $\depth$			\\
Local wavespeed		& $c = \sqrt{g \depth}$	\\
Local peak wavelength	& $\lam = \sqrt{gd}/f_p$	\\
Depth ratio			& $\drat = \depth/\dup$	
\end{tabular}
\end{center}
\end{table}
 %^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%

We introduce the following characteristic scales
\begin{align}
\label{ampdefn}
&\amp = \mean{\eta^2}^{1/2} 
&&\mbox{\em characteristic wave amplitude} \\
\label{lamdefn}
&\lam = \sqrt{g \depth} / \freqp
&&\mbox{\em characteristic wave length}
\end{align}
where $\mean{\cdot}$ indicates the mean of a quantity. Note that the characteristic wavelength $\lam = \lam(x)$ takes different values upstream and downstream of the ADC. We remark that experimental measurements indicate that the characteristic amplitude $\amp = \mean{\eta^2}^{1/2}$ is nearly the same on both values of the ADC. Hence, we will not distinguish between upstream and downstream values of $\amp$.
We introduce the following dimensionless variables
\begin{align}
&u = \eta/\amp
&&\mbox{\em dimensionless surface displacement} \\
&\tilde{x} = \frac{2 \pi \xi}{\lamfac \lam} = \frac{2 \pi (x-ct)}{\lamfac \lam}
&&\mbox{\em dimensionless position in moving frame} \\
& \tilde{t} = t \freqp / \lamfac
&&\mbox{\em dimensionless time} \\
\end{align}
We will consider the physical domain $\xi \in [0, \lamfac \lam]$, which is an integer multiple of the characteristic wavelength where $\lamfac$ will be selected later (probably about 5). Above, we have chosen the scale to normalize position as $\lamfac \lam/(2\pi)$ so that the dimensionless domain can be taken (after a shift) to be $\tilde{x} \in [-\pi,\pi]$. Accordingly, we have also chosen the time scale as $\lamfac/\freqp$, so that a wave traveling at the local wavespeed $c = \lam \freqp = \sqrt{g \depth}$ crosses the physical domain $x \in [0, \lamfac \lam]$ in one unit of dimensionless time. Said differently, the dimensionless wave speed is $(2 \pi)^{-1}$ so that a wave crosses the domain $[-\pi,\pi]$ in unit time.

Dropping the tildes, the dimensionless KdV equation takes the form
\begin{equation}
\label{dimlessKdV}
u_t + {3 \pi} \epsup \drat^{-1} \, u u_x + \frac{4 \pi^3}{3 \lamfac^2} \delup^2 \drat \, u_{xxx} = 0
\qquad \text{for } x \in [-\pi,\pi]
\end{equation}
where dimensionless variables are as follows
\begin{align}
&\epsup = \amp/\dup \, , \\
&\delup = \dup/\lamup = \sqrt{\dup \freqp^2/g} \, \\
&\drat(x) = {\depth(x)}/{\dup} \, .
\end{align}
Here, $\drat = \drat(x)$ is the local dimensionless depth. See Table \ref{paramtable} for a summary of important parameters.

@@

Then the Hamiltonian of \eqref{dimlessKdV} is given by
\begin{equation}
\Ham = 3 \pi \epsup \drat^{-1} \, H_3 - \frac{4 \pi^3}{3 \lamfac^2} \delup^2 \drat \, H_2
\end{equation}

\begin{comment}
Old Note (around May 2019): After giving it some deliberation, I believe we should use $\lamfac=1$. My reason is that the peak in the wave spectrum occurs at a frequency of $\freqp$ or wavelength of $\lam$. In the tKdV Gibbs measure, the spectrum decays monotonically, so that the peak is at the lowest resolved frequency (or largest resolved wavelength). Thus, we want $\lam$ to correspond to the largest resolved wavelength, i.e.~the length of the periodic domain in the tKdV framework.

New Thought (July 2019): Actually, it is very possible that the bandwidth (2 Hz in the experiments) sets the value of $\lamfac$ that is most appropriate, since that bandwidth sets the decay rate of the spectrum. My thinking is that our experimental forcing does not really follow the upstream Gibbs measure all that well (because we did not have that in mind), but perhaps it best approximates the tail decay of some Gibbs measure. Perhaps, the lowest frequencies (i.e. those much slower than 2 Hz that are present in the theory but almost absent in the experiments) do not affect statistics that much.
\end{comment}

\subsection{Matching condition}

The statistical matching condition is
\begin{equation}
\meanup{\Hp} = \meandn{\Hp}
\end{equation}
An experimental observation is that the incoming skewness is small, thus
\begin{equation}
\meanup{H_3} \approx 0
\end{equation}
which immediately gives
\begin{equation}
\ampp D^{-3/2} \meandn{H_3} = \mu D^{1/2} \left( \meandn{H_2} - \meanup{H_2} \right)
\end{equation}
or equivalently
\begin{equation}
\label{H3H2}
\meandn{H_3} = \frac{\mu D^{2}}{\ampp} \, \Delta \mean{H_2}
\end{equation}
where $\Delta \mean{H_2} =  \meandn{H_2} - \meanup{H_2}$  indicates the difference between the upstream and downstream values.

>>>>

% Basic physical parameters and scales
\newcommand{\depth}{h}
\newcommand{\dup}{\depth_{-}}
\newcommand{\ddn}{\depth_{+}}
\newcommand{\freqp}{f_p}
\newcommand{\lam}{\lambda}
\newcommand{\lamup}{\lam_{-}}
\newcommand{\lamfac}{N}
\newcommand{\amp}{a}
%\newcommand{\etastd}{\eta_{\text std}}
% Statistical quantities
\newcommand{\skw}{\text{skew}}
\newcommand{\var}{\text{var}}
% Dimensionless parameters
\newcommand{\epsup}{\eps_0}
\newcommand{\delup}{\delta_0}
\newcommand{\drat}{D}
\newcommand{\dratdn}{\drat_0}
\newcommand{\ampp}{\mathcal{A}}
% Hamiltonian and Gibbs stuff
\newcommand{\Ham}{H}
\newcommand{\Hup}{\Ham^{-}}
\newcommand{\Hdn}{\Ham^{+}}
\newcommand{\Gibbs}{\mathcal{G}}
\newcommand{\Gup}{\Gibbs^{-}}
\newcommand{\Gdn}{\Gibbs^{+}}
\newcommand{\meanup}[1]{\mean{#1}_{-}}
\newcommand{\meandn}[1]{\mean{#1}_{+}}

@@

\author{
C. Tyler Bolles\thanks{University of Michigan},
Andrew J.~Majda\thanks{Courant Institute of Mathematical Sciences}, 
M.~N.~J.~Moore\thanks{Florida State University}, 
Di Qi\thanks{Courant Institute of Mathematical Sciences} }
\maketitle

@@

We will primarily consider the case in which waves originate from a region of constant depth, encounter an abrupt depth change, and then continue in another region of constant depth. Thus, depth will be piecewise constant
\begin{align}
\depth = 
\begin{cases}
\dup \quad \mbox{if } x<0 \\
\ddn \quad \mbox{if } x>0
\end{cases}
\end{align}
Most often, we will consider waves moving into shallower depth, so that $\dup > \ddn$. The incoming waves are randomized and generated with a peak forcing frequency of $\freqp$, which gives rise to peak wavelength $\lam = c/\freqp = \sqrt{g \depth} / \freqp$.

@@

% Table
%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
\begin{table}[h]%[htbp]
\begin{center}
\caption{Table of parameters}
\label{paramtable}
\begin{tabular}{l l l}
\hline \multicolumn{3} { c }{Parameters that are constant in a single experiment} \\
\hline Description & Notation and definition & Value in experiments \\
\hline
Peak forcing frequency		& $f_p$						& 2 Hz \\
Upstream depth			& $\dup$						& 12.5 cm \\
Downstream depth			& $\ddn$						& 3 cm \\
Characteristic wave amplitude	& $a = \mean{\eta^2}^{1/2} $		& 0.03--0.3 cm \\
Upstream wavelength		& $\lamup = \sqrt{g \dup}/f_p$		& 56 cm \\
Amplitude-to-depth ratio		& $\epsup = a / \dup$			& 0.0024--0.024 \\
Depth-to-wavelength ratio		& $\delup = \dup / \lamup$		& 0.22 \\
Depth ratio				& $\dratdn = \ddn/\dup$			& 0.24 \\
\hline \multicolumn{3} { c }{Parameters that change value across ADC} \\
\hline Description & Notation and definition \\
\hline
Local depth			& $\depth$			\\
Local wavespeed		& $c = \sqrt{g \depth}$	\\
Local peak wavelength	& $\lam = \sqrt{gd}/f_p$	\\
Local dimensionless depth		& $\drat = \depth/\dup$	
\end{tabular}
\end{center}
\end{table}
 %^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%

We introduce the following characteristic scales
\begin{align}
\label{ampdefn}
&\amp = \mean{\eta^2}^{1/2} 
&&\mbox{\em characteristic wave amplitude} \\
\label{lamdefn}
&\lam = \sqrt{g \depth} / \freqp
&&\mbox{\em characteristic wave length}
\end{align}
where $\mean{\cdot}$ indicates the mean of a quantity. Note that the characteristic wavelength $\lam = \lam(x)$ takes different values upstream and downstream of the ADC. We remark that experimental measurements indicate that the characteristic amplitude $\amp = \mean{\eta^2}^{1/2}$ is nearly the same on both values of the ADC. Hence, we will not distinguish between upstream and downstream values of $\amp$.
We introduce the following dimensionless variables
\begin{align}
&u = \eta/\amp
&&\mbox{\em dimensionless surface displacement} \\
&\tilde{x} = \frac{2 \pi \xi}{\lamfac \lam} = \frac{2 \pi (x-ct)}{\lamfac \lam}
&&\mbox{\em dimensionless position in moving frame} \\
& \tilde{t} = t \freqp / \lamfac
&&\mbox{\em dimensionless time}
\end{align}
We will consider the physical domain $\xi \in [0, \lamfac \lam]$, which is an integer multiple of the characteristic wavelength where $\lamfac$ will be selected later (probably about 5). Above, we have chosen the scale to normalize position as $\lamfac \lam/(2\pi)$ so that the dimensionless domain can be taken (after a shift) to be $\tilde{x} \in [-\pi,\pi]$. Accordingly, we have also chosen the time scale as $\lamfac/\freqp$, so that a wave traveling at the local wavespeed $c = \lam \freqp = \sqrt{g \depth}$ crosses the physical domain $x \in [0, \lamfac \lam]$ in one unit of dimensionless time. Said differently, the dimensionless wave speed is $(2 \pi)^{-1}$ so that a wave crosses the domain $[-\pi,\pi]$ in unit time.

%----------------------------------------------------------------------------%
% COMMENT ON LAMFAC
\begin{comment}
Old Note (around May 2019): After giving it some deliberation, I believe we should use $\lamfac=1$. My reason is that the peak in the wave spectrum occurs at a frequency of $\freqp$ or wavelength of $\lam$. In the tKdV Gibbs measure, the spectrum decays monotonically, so that the peak is at the lowest resolved frequency (or largest resolved wavelength). Thus, we want $\lam$ to correspond to the largest resolved wavelength, i.e.~the length of the periodic domain in the tKdV framework.

New Thought (July 2019): Actually, it is very possible that the bandwidth (2 Hz in the experiments) sets the value of $\lamfac$ that is most appropriate, since that bandwidth sets the decay rate of the spectrum. My thinking is that our experimental forcing does not really follow the upstream Gibbs measure all that well (because we did not have that in mind), but perhaps it best approximates the tail decay of some Gibbs measure. Perhaps, the lowest frequencies (i.e. those much slower than 2 Hz that are present in the theory but almost absent in the experiments) do not affect statistics that much.
\end{comment}
%----------------------------------------------------------------------------%

Dropping the tildes, the dimensionless KdV equation takes the form
\begin{equation}
\label{dimlessKdV}
u_t + {3 \pi} \epsup \drat^{-1} \, u u_x + \frac{4 \pi^3}{3 \lamfac^2} \delup^2 \drat \, u_{xxx} = 0
\qquad \text{for } x \in [-\pi,\pi]
\end{equation}
where dimensionless variables are as follows
\begin{align}
&\epsup = \amp/\dup \, , \\
&\delup = \dup/\lamup = \sqrt{\dup \freqp^2/g} \, \\
&\drat = {\depth}/{\dup} \, .
\end{align}
Here, $\drat$ is the local dimensionless depth, which changes value crossing the ADC. In particular, since \eqref{dimlessKdV} is in the reference frame that moves with the characteristic wave speed, we can assume that the ADC is met at $t = T_{ADC} = 0$. Then we have
\begin{equation}
\drat = 
\begin{cases}
1 		&\quad \mbox{for } t<0 \\
\dratdn 	&\quad \mbox{for } t>0
\end{cases}
\end{equation}
where $\dratdn = \ddn/\dup$.
See Table \ref{paramtable} for a summary of important parameters.

@@

Then the Hamiltonian of \eqref{dimlessKdV} is given by
\begin{equation}
\Ham = 3 \pi \epsup \drat^{-1} \, H_3 - \frac{4 \pi^3}{3 \lamfac^2} \delup^2 \drat \, H_2
\end{equation}
where $\drat$ changes value across the ADC. More specifically, there exists an upstream and a downstream Hamiltonian
\begin{align}
&\Hup = 3 \pi \epsup \, H_3 - \frac{4 \pi^3}{3 \lamfac^2} \delup^2 \, H_2 \\
&\Hdn = 3 \pi \epsup \dratdn^{-1} \, H_3 - \frac{4 \pi^3}{3 \lamfac^2} \delup^2 \dratdn \, H_2
\end{align}
$\Hup$ is a conserved quantity for $t<0$ and $\Hdn$ is a conserved quantity for $t>0$. Each Hamiltonian $\Ham^{\pm}$generates a corresponding Gibbs measure $\Gibbs^{\pm}$. (FILL IN DEFINITIONS)

\vsp{10} 
\subsection{Matching condition}

The statistical matching condition is
\begin{equation}
\meanup{\Hdn} = \meandn{\Hdn}
\end{equation}
An experimental observation is that the incoming skewness is small, thus
\begin{equation}
\meanup{H_3} \approx 0
\end{equation}
which immediately gives
\begin{equation}
\ampp D^{-3/2} \meandn{H_3} = \mu D^{1/2} \left( \meandn{H_2} - \meanup{H_2} \right)
\end{equation}
or equivalently
\begin{equation}
\label{H3H2}
\meandn{H_3} = \frac{\mu D^{2}}{\ampp} \, \Delta \mean{H_2}
\end{equation}
where $\Delta \mean{H_2} =  \meandn{H_2} - \meanup{H_2}$  indicates the difference between the upstream and downstream values.
